Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans

### 2020-05-29 paper_PLOS/paper_PLOS_draft.tex
Figure 2 corrected

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

\hypertarget{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}{%
\subsection*{Step-wise reduction of intrinsic cancer risk in large,
long-lived
Afrotherians}\label{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}}
\addcontentsline{toc}{subsection}{Step-wise reduction of intrinsic
cancer risk in large, long-lived Afrotherians}

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

\hypertarget{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}{%
\subsection{Step-wise reduction of intrinsic cancer risk in large,
long-lived
Afrotherians}\label{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}}

\includegraphics[width=6in]{paper_PLOS_draft_files/figure-latex/Figure2-1} \hfill{}

\caption{Cancer succecptibility across Atlantogenata}\label{fig:Figure2}
\end{figure}

\begin{longtable}[]{@{}lrlllr@{}}
\caption{Estimated Cancer Succeptibilty for nodes in
Atlantogenata}\tabularnewline
\toprule
Node & Est. Lifespan & K1 & K2 & Change in K & log2
Change\tabularnewline
\midrule
\endfirsthead
\toprule
Node & Est. Lifespan & K1 & K2 & Change in K & log2
Change\tabularnewline
\midrule
\endhead
Loxodontini & 34.38 & 1.47e+16 & 2.97e+15 & 4.94e+00 &
2.31\tabularnewline
Loxodonta africana & 65.00 & 2.47e+17 & 1.47e+16 & 1.68e+01 &
4.07\tabularnewline
Loxodona & 34.38 & 1.47e+16 & 1.47e+16 & 1.00e+00 & 0.00\tabularnewline
Loxodonta cyclotis & 31.12 & 2.97e+15 & 1.47e+16 & 2.02e-01 &
-2.31\tabularnewline
Palaeoloxodon antiquus & 34.38 & 1.47e+16 & 1.47e+16 & 1.00e+00 &
0.00\tabularnewline
Elephantidae & 31.12 & 2.97e+15 & 1.40e+07 & 2.13e+08 &
27.66\tabularnewline
Elephantina & 34.38 & 1.47e+16 & 2.97e+15 & 4.94e+00 &
2.31\tabularnewline
Elephas maximus & 65.50 & 2.58e+17 & 1.47e+16 & 1.76e+01 &
4.14\tabularnewline
Mammuthus & 34.38 & 1.47e+16 & 1.47e+16 & 1.00e+00 & 0.00\tabularnewline
Mammuthus primigenius & 31.12 & 2.97e+15 & 1.47e+16 & 2.02e-01 &
-2.31\tabularnewline
Mammuthus columbi & 34.38 & 1.47e+16 & 1.47e+16 & 1.00e+00 &
0.00\tabularnewline
Proboscidea & 9.41 & 1.40e+07 & 1.21e+14 & 1.15e-07 &
-23.05\tabularnewline
Mammut americanum & 34.38 & 1.47e+16 & 1.40e+07 & 1.05e+09 &
29.97\tabularnewline
Tethytheria & 25.49 & 1.21e+14 & 1.01e+12 & 1.21e+02 &
6.92\tabularnewline
Trichechus manatus & 69.00 & 4.77e+16 & 1.21e+14 & 3.93e+02 &
8.62\tabularnewline
Paenungulata & 18.91 & 1.01e+12 & 1.01e+12 & 1.00e+00 &
0.00\tabularnewline
Procavia capensis & 14.80 & 3.13e+10 & 1.01e+12 & 3.11e-02 &
-5.01\tabularnewline
Pseudoungulata & 18.91 & 1.01e+12 & 1.69e+09 & 5.97e+02 &
9.22\tabularnewline
Orycteropus afer & 29.80 & 4.19e+13 & 1.01e+12 & 4.17e+01 &
5.38\tabularnewline
Elephantulus edwardii & 10.40 & 6.90e+07 & 1.69e+09 & 4.09e-02 &
-4.61\tabularnewline
Afrosoricida & 10.40 & 6.90e+07 & 1.69e+09 & 4.09e-02 &
-4.61\tabularnewline
Chrysochloris asiatica & 10.40 & 6.90e+07 & 6.90e+07 & 1.00e+00 &
0.00\tabularnewline
Echinops telfairi & 19.00 & 2.57e+09 & 6.90e+07 & 3.72e+01 &
5.22\tabularnewline
Afrotheria & 12.69 & 1.69e+09 & 2.83e+06 & 5.97e+02 &
9.22\tabularnewline
Xenarthra & 20.89 & 4.97e+12 & 2.83e+06 & 1.76e+06 &
20.75\tabularnewline
Dasypus novemcinctus & 22.30 & 3.67e+11 & 4.97e+12 & 7.37e-02 &
-3.76\tabularnewline
Choloepus hoffmanni & 41.00 & 1.42e+13 & 4.97e+12 & 2.85e+00 &
1.51\tabularnewline
Atlantogenata & 8.52 & 2.83e+06 & 2.83e+06 & 1.00e+00 &
0.00\tabularnewline
Afroinsectivora & 12.69 & 1.69e+09 & 1.69e+09 & 1.00e+00 &
0.00\tabularnewline
\bottomrule
\end{longtable}

The dramatic increase in body mass and lifespan in some Afrotherian
lineages implies those lineages evolved reduced cancer risk. To infer
the magnitude of these reductions we estimated differences in cancer
risk between small bodied, short-lived species and large bodied,
long-lived species as well as for reconstructed ancestral Afrotherians.
Following {[}44{]} we estimate the intrinsic cancer risk as the product
of risk associated with body mass and lifespan. Differences in cancer
susceptibility \(K\) due to body mass differences between species can be
approximated simply as the fold difference in body mass (\(D\)) between
species {[}44{]}. The risk of developing cancer also increases in
proportion to the sixth power of age and is approximated by the formula
\(Ct^6\), in which the proportionality constant C that determines
susceptibility to cancer induction is multiplied by the sixth power of
the age in years, \(t\) {[}44--46{]}. Thus we can estimate the intrinsic
cancer risk for a species as \(K \approx Dt^6\).

In order to estimate the intrinsic cancer risk of a species, we first
obtained estimates for lifespans at ancestral nodes using PGLS and the
model
\(ln(lifespan) = \beta_{1}corBrownian +\beta_{2}ln(Size) + \epsilon\)
(Table 2). With this information in hand, we calculated \(K_{1}\) at all
nodes, and then estimated the fold change in cancer susceptibility
between an ancestral node and a given node as \(\frac{K_{2}}{K_{1}}\)
(\textbf{Table 4}). As body size and phylogeny are the strongest
predictors of lifespan, we find that this regression is sufficiently
robust without further metabolic or other covariates for our purposes
herein.

As shown in \textbf{Table 2}, cancer susceptibility skyrocketed at the
initial divergence of Atlantogenata, followed by a generally upwards
trend. At the common ancestor of Afrotheria there is an initial
9.22-fold increase in cancer risk. In parallel to Afrotheria, cancer
susceptibility increases 20.75-fold in Xenarthra. However, cancer risk
slowly deflates as size decreases as one moves along the tree towards
extant species, such as in Hoffman's Two Toed Sloth (1.51-fold change)
and in the Nine-banded Armadillo (-3.76-fold change).

Within Afrotheria, cancer susceptibility drops in Afrosoricida as
species shrink (-4.61-fold, then stagnates for the Cape Golden Mole) -
but then rises 5.22-fold towards the Lesser Hedgehog Tenrec. In
parallel, Afroinsectivora does not increase in cancer susceptibility,
and decreases once more at the Cape Elephant Shrew (-4.61-fold). The
emergence of Pseudoungulata sees the next big leap in cancer
susceptibility with a 9.22-fold increase. The Aardvark further increases
5.38-fold, while we don't observe an increase at the common ancestor of
Paenungulates. While the Rock Hyrax decreases in cancer susceptibility
as expected (-5.01-fold), Tethytheria sees a sharp increase in cancer
risk (6.92-fold). Within Tethytheria, the Manatee's cancer risk
increases once more 8.62-fold, while Proboscidea's cancer risk drops
precipitously along with its body size (-23.05-fold). Yet, within
Proboscidea we see the biggest increases: right off the bat, we see that
the cancer susceptibility of Elephantidae and the American Mastodon
skyrocket by 27.66-fold and 29.97-fold, respectively. Both Elephantina
and Loxodontini in Elephantidae have a 2.31-fold increase in cancer
susceptibility. Within Elephantina, cancer susceptibility stays stable
at Mammuthus and in the Colombian Mammoth, and slightly decreases in the
Wooly Mammoth (-2.31-fold). The three extant elephants - Asian Elephant
in Elephantina, the African Savana Elephant in Loxodontini, and the
African Forest Elephant in Loxodona, meanwhile, have parallel and
similar decreases in both size and cancer susceptibility (4.14-, 4.07-,
and -2.31-fold, respectively). Neither the common ancestor of Loxodonta,
nor the Straight-Tusked Mammoth see any further changes in cancer
susceptibility.

\hypertarget{identification-and-evolutionary-history-of-gene-duplications}{%
\subsection*{Identification and evolutionary history of gene
duplications}\label{identification-and-evolutionary-history-of-gene-duplications}}
\addcontentsline{toc}{subsection}{Identification and evolutionary
history of gene duplications}

@@

\leavevmode\hypertarget{ref-ArmitageAndDoll1954}{}%
45. Armitage P, Doll R. The Age Distribution of Cancer and a Multi-stage
Theory of Carcinogenesis. British Journal of Cancer. 1954;8: 1--12.
doi:\href{https://doi.org/10.1038/bjc.1954.1}{10.1038/bjc.1954.1}

\leavevmode\hypertarget{ref-Nordling1953}{}%
46. Nordling CO. A New Theory on the Cancer-inducing Mechanism. British
Journal of Cancer. 1953;7: 68--72.
doi:\href{https://doi.org/10.1038/bjc.1953.8}{10.1038/bjc.1953.8}
